---
title: "Real Data Analysis: Riboflavin Dataset"
subtitle: "Comparing RIGHT against other robust and standard methods"
format: html
editor: visual
---

## 1. Data Loading and Preprocessing

In this section, we load the `riboflavin` dataset and perform a robust preprocessing routine to prepare it for analysis.

```{r data-preprocessing}
library(dplyr)
library(glmnet)
library(hdi)
library(ggplot2)
source("../../R/solvers.R")
data(riboflavin)

# Assign to more standard names
X_raw <- riboflavin$x
y_raw <- riboflavin$y
#summary(X_raw[, 1:5])
cat(sprintf("Dataset loaded: %d samples, %d features.\n", nrow(X_raw), ncol(X_raw)))

# --- 2. Robust Preprocessing ---
# We use a robust standardization procedure for all methods to ensure a fair
# comparison while preserving the heavy-tailed nature of the data.

# Center the response variable by its median
y_centered <- y_raw - median(y_raw)

# Robustly standardize the feature matrix
# We define a helper function for this
robust_scale <- function(m) {
  # Center each column by its median
  m_centered <- apply(m, 2, function(col) col - median(col))
  
  # Scale each column by its Median Absolute Deviation (MAD)
  # for normally distributed data. We add a small epsilon for stability.
  mads <- apply(m_centered, 2, mad)
  m_scaled <- sweep(m_centered, 2, mads, FUN = "/")
  
  # Handle columns with zero MAD (constant columns) by setting them to zero
  m_scaled[ , mads == 0] <- 0
  
  return(m_scaled)
}

X_scaled <- robust_scale(X_raw)

# --- 3. Final Data Objects ---
# These will be the inputs for all subsequent analyses
X <- X_scaled
y <- y_centered

cat("\nData preprocessing complete.\n")
# Display a summary of the first 5 columns to show the effect of scaling
summary(X[, 1:5])
```

## 2. Exploratory Data Analysis: Assessing Heavy Tails

Before comparing the methods, we first investigate the distributional properties of the (raw) response and the features to justify the need for robust estimators.

```{r}
#| echo: true
# We use the raw data objects X_raw and y_raw loaded in the first chunk.
library(moments)
library(ggplot2)

# --- Analysis of Raw Response y_raw ---
y_raw_df <- data.frame(y = y_raw)
p_y_raw_qq <- ggplot(y_raw_df, aes(sample = y)) +
  stat_qq(color = "red", size = 2.5) +
  stat_qq_line(linewidth = 1.2) +
  labs(
    x = "Theoretical Quantiles (Normal)",
    y = "Sample Quantiles (response)"
  ) +
  theme_minimal(base_size = 20)

print(p_y_raw_qq)
ggsave("../../results/06_Sixth_real_data/real_data_y_qq.pdf",p_y_raw_qq,height = 4.5,width=10,dpi=300)
ggsave("../../results/06_Sixth_real_data/real_data_y_qq.png",p_y_raw_qq,height = 4.5,width=10,dpi=300)

# --- Analysis of Raw Features X_raw ---
cat("Summary of Raw Feature Kurtosis (X_raw):\n")
raw_kurtosis <- apply(X_raw, 2, kurtosis)
summary(raw_kurtosis)

kurtosis_raw_df <- data.frame(kurtosis = raw_kurtosis)
p_x_raw_kurt <- ggplot(kurtosis_raw_df, aes(x = kurtosis)) +
  geom_histogram(bins = 50, fill = "skyblue", color = "black") +
  geom_vline(xintercept = 3, color = "blue", linetype = "dashed", linewidth = 1.2) +
  labs(
    #title = "Distribution of Kurtosis for Raw Features (X_raw)",
    x = "Kurtosis",
    y = "Number of Features"
  ) +
  scale_x_log10() +
  theme_minimal(base_size = 20)

print(p_x_raw_kurt)
ggsave("../../results/06_Sixth_real_data/real_data_x_kurt.pdf",p_x_raw_kurt,height = 4.5,width=10,dpi=300)
ggsave("../../results/06_Sixth_real_data/real_data_x_kurt.png",p_x_raw_kurt,height = 4.5,width=10,dpi=300)

summary(y)
summary(X[,1:5])

```

## 3. Model Training and Hyperparameter Tuning

In this section, we train each model and tune its hyperparameters. We begin by splitting the data into a single, fixed training (80%) and test (20%) set. All hyperparameter tuning is performed exclusively on the training set.

```{r model-tuning-and-fitting}
# --- 1. Create a Single, Fixed Data Split ---
set.seed(42) # for reproducibility of the split
train_proportion <- 0.8
train_idx <- sample(1:nrow(X), floor(train_proportion * nrow(X)))

X_train <- X[train_idx, ]
y_train <- y[train_idx]
X_test <- X[-train_idx, ]
y_test <- y[-train_idx]

cat(sprintf("Data split into %d training samples and %d test samples.\n", nrow(X_train), nrow(X_test)))

# --- Initialize a list to store the final trained models ---
final_models <- list()

# ==============================================================================
# STAGE 1: TUNE INDEPENDENT METHODS
# ==============================================================================
cat("\n--- Stage 1: Tuning Lasso, Shrinkage, and Adaptive Huber ---\n")

# --- 1a. Tune and Fit Lasso ---
cat("Tuning Lasso...\n")
cv_fit_lasso <- cv.glmnet(X_train, y_train, standardize = FALSE, intercept = FALSE, nfolds = 10)
lambda_best_lasso <- cv_fit_lasso$lambda.min
final_models$Lasso <- as.vector(coef(cv_fit_lasso, s = lambda_best_lasso))[-1]

# --- 1b. Tune and Fit Shrinkage ---
cat("Tuning Shrinkage...\n")
tau_x_shrinkage <- quantile(abs(X_train), 0.95)
# y is only centered, not scaled, so tau_y is based on centered y_train
tau_y_shrinkage <- quantile(abs(y_train), 0.95) 
X_train_shrunk <- truncate_operator(X_train, tau_x_shrinkage)
y_train_shrunk <- truncate_operator(y_train, tau_y_shrinkage)
cv_fit_shrinkage <- cv.glmnet(X_train_shrunk, y_train_shrunk, standardize = FALSE, intercept = FALSE, nfolds = 10)
lambda_best_shrinkage <- cv_fit_shrinkage$lambda.min
final_models$Shrinkage <- solver_shrinkage_method(X_train, y_train, tau_x_shrinkage, tau_y_shrinkage, lambda_best_shrinkage)

# --- 1c. Tune and Fit Adaptive Huber ---
cat("Tuning Adaptive Huber...\n")
# Step 1: Get initial residuals from the tuned Lasso on the training set
residuals_init <- y_train - X_train %*% final_models$Lasso

tau_best_huber <- quantile(abs(residuals_init), 0.95)

# Step 2: Tune lambda for Huber using a k-fold CV loop on the TRAINING data
k_folds <- 4
fold_indices <- sample(cut(seq(1, nrow(X_train)), breaks = k_folds, labels = FALSE))
lambda_grid_huber <- 10^seq(-2, 1, length.out = 20)
cv_errors_huber <- sapply(lambda_grid_huber, function(lambda_val) {
  fold_mses <- sapply(1:k_folds, function(k) {
    val_fold_idx <- which(fold_indices == k)
    train_fold_idx <- which(fold_indices != k)
    
    theta_tuned <- solver_huber_lasso(X_train[train_fold_idx, ], y_train[train_fold_idx], 
                                      lambda_val, tau_best_huber, T_max = 250)
    
    y_pred_val <- X_train[val_fold_idx, ] %*% theta_tuned
    mean((y_train[val_fold_idx] - y_pred_val)^2)
  })
  mean(fold_mses)
})
lambda_best_huber <- lambda_grid_huber[which.min(cv_errors_huber)]

# Final fit of Adaptive Huber on the full training set with best params
final_models$`Adaptive Huber` <- solver_huber_lasso(X_train, y_train, lambda_best_huber, tau_best_huber, T_max = 500)

# ==============================================================================
# STAGE 2: DETERMINE SPARSITY AND FIT DEPENDENT METHODS
# ==============================================================================
cat("\n--- Stage 2: Determining Sparsity and Fitting RIGHT & IHT ---\n")

# --- 2a. Determine Sparsity 's' from the final Adaptive Huber model ---
s_from_huber <- sum(final_models$`Adaptive Huber` != 0)
# Apply a reasonable cap (e.g., max 1/4 of training samples) to avoid overly dense models
s_from_lasso<-sum(final_models$`Lasso` != 0)
s_final <- s_from_lasso
cat(sprintf("Sparsity 's' derived from Adaptive Huber model: %d\n", s_final))

# --- 2b. Fit RIGHT ---
cat("Fitting RIGHT...\n")
K_right <- max(1, round( 5/9*log(ncol(X_train) * log(nrow(X_train) / log(ncol(X_train))))))
final_models$RIGHT <- solver_right(X_train, y_train, s =s_final, eta = 0.01, T_max = 250, 
                                   K = K_right, grad_func_samplewise = grad_linear_regression_samplewise)

# --- 2c. Fit IHT ---
cat("Fitting IHT...\n")
final_models$IHT <- solver_iht(X_train, y_train, s = s_final, eta = 0.01, T_max = 250, 
                               grad_func = grad_linear_regression)

cat("\n--- All models have been trained on the training data. ---\n")
saveRDS(final_models,"../../results/06_Sixth_real_data/final_models.rds")
```

## 4. Performance Evaluation on Test Set

Now that all models are trained and their hyperparameters are tuned, we evaluate their predictive performance on the held-out test set `(X_test, y_test)`. We use Mean Absolute Error (MAE) as our primary metric, as it is robust to potential outliers in the test set responses.\`

```{r}
# --- 1. Calculate MAE for each model ---
mae_results <- sapply(final_models, function(beta_hat) {
  # Handle cases where a solver might have failed and returned NA
  if (any(is.na(beta_hat))) {
    return(NA_real_)
  }
  
  # Predict on the test set
  y_pred <- X_test %*% beta_hat
  
  # Calculate Mean Absolute Error
  mae <- mean(abs(y_test - y_pred))
  
  return(mae)
})



# --- 2. Format and Display Results ---
# Convert to a data frame for nice printing
results_df <- data.frame(
  Method = names(mae_results),
  MAE = mae_results
) %>%
  arrange(MAE) # Sort by MAE to see the best performing method on top

cat("--- Predictive Performance on Test Set ---\n")
print(results_df)

# --- 3. Visualize the Comparison ---
p_mae <- ggplot(results_df, aes(x = reorder(Method, MAE), y = MAE, fill = Method)) +
  geom_col() +
  geom_text(aes(label = round(MAE, 4)), vjust = -0.3, size = 4) +
  labs(
    title = "Model Performance on Riboflavin Test Set",
    x = "Method",
    y = "Mean Absolute Error (MAE)"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none") # No legend needed as x-axis is clear

print(p_mae)

cat("shrinkage_s:",sum(final_models$`Shrinkage` != 0),"\n")
cat("Adaptive Huber_s",sum(final_models$`Adaptive Huber` != 0),"\n")
cat("IHT_s",sum(final_models$`IHT` != 0),"\n")
cat("RIGHT_s",sum(final_models$`RIGHT` != 0),"\n")
cat("lasso_s",sum(final_models$`Lasso` != 0),"\n")

```

### 4.1 Residual Analysis for the Best Model (RIGHT)

To further understand the behavior of our best-performing model, we analyze the distribution of its residuals on the test set. A Q-Q plot helps us visualize whether the residuals follow a normal distribution or exhibit heavy tails.

```{r residual-analysis}
# --- 1. Calculate the Residuals for the RIGHT model ---
# Get the final trained RIGHT model from our list
beta_hat_right <- final_models$RIGHT

# Predict on the test set
y_pred_right <- X %*% beta_hat_right

# Calculate the residuals
residuals_right <- y - y_pred_right

# Create a data frame for plotting
residuals_df <- data.frame(residuals = residuals_right)

# --- 2. Create the Q-Q Plot ---
p_residuals_qq <- ggplot(residuals_df, aes(sample = residuals)) +
  stat_qq(color = "darkcyan", size = 2.5) +
  stat_qq_line(linewidth = 1.2, color = "black") +
  labs(
    #title = "Q-Q Plot of Residuals for the RIGHT Model",
    #subtitle = "Comparison against a Normal Distribution on the Test Set",
    x = "Theoretical Quantiles (Normal)",
    y = "Sample Quantiles (Residuals)"
  ) +
  theme_minimal(base_size = 20)

print(p_residuals_qq)
ggsave("../../results/06_Sixth_real_data/real_data_residuals_qq.pdf",p_residuals_qq,height = 4.5,width=10,dpi=300)
ggsave("../../results/06_Sixth_real_data/real_data_residuals_qq.png",p_residuals_qq,height = 4.5,width=10,dpi=300)

# --- 3. (Optional) Formal Test for Normality ---
# The Shapiro-Wilk test is a formal test for normality.
# The null hypothesis is that the data is normally distributed.
# A small p-value (< 0.05) suggests we can reject the null hypothesis.
shapiro_test_result <- shapiro.test(residuals_right)
cat("\n--- Shapiro-Wilk Normality Test on Residuals ---\n")
print(shapiro_test_result)
```

计算MSE

```{r}
mse_results <- sapply(final_models, function(beta_hat) {
  # Handle cases where a solver might have failed and returned NA
  if (any(is.na(beta_hat))) {
    return(NA_real_)
  }
  
  # Predict on the test set
  y_pred <- X_test %*% beta_hat
  
  # Calculate Mean Absolute Error
  mse <- mean((y_test - y_pred)^2)
  
  return(mse)
})

# --- 2. Format and Display Results ---
# Convert to a data frame for nice printing
results_df_mse <- data.frame(
  Method = names(mse_results),
  MSE = mse_results
) %>%
  arrange(MSE) # Sort by MAE to see the best performing method on top

cat("--- Predictive Performance on Test Set ---\n")
print(results_df_mse)

```
